Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model

### 2025-10-09 paper/paper.qmd
describe models in results

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  insert_loa: false
---

@@

![System dynamics model](images/sd_model.svg){#fig-sd-model}

@@

A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the exception that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak. The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate). The parameter $r$ (rate of random infection) was kept constant throughout the simulation.

@@

@@

Rejection sampling \acr{ABC} was employed to fit the model to the outbreak data [@sunnakerApproximateBayesianComputation2013]. Priors were extracted from existing literature; flat priors when data were unavailable. A summary of the priors used is available in @tbl-priors. The proportion of toilets that were shared between multiple patients was set to 60%.

@@

As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates. Three approaches to sampling from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were used:

1. Parameter samples drawn from a uniform distribution:

$$
U_[a,b] = U(\mu - \sigma, \mu + \sigma),
$$

where: $\mu$ and $\sigma$ are the mean and standard deviation of the posterior distribution, respectively;

2. The mean of each parameter's posterior distribution; and
3. Samples with replacement taken from the posterior distribution.

@@

One of the simulation runs was compared to the deterministic model, which was numerically integrated using Euler's method.

![Screenshot of model graphical user interface during a simulation run](./images/screenshot.png){#fig-screenshot height="7in"}

Users can run the model interactively using the interface shown in @fig-screenshot.

@@

To assess the sensitivity of the calibrated model to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

An \acr{ABM} that simulates the faecal-oral transmission of infection or colonisation in a hospital setting was produced. After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

@@

```{r fig-consts-sims-mean-sd, include=TRUE}
#| echo: false
#| fig-cap: "Simulated rates of _C. difficile_ infection generated by using the posterior parameter distributions. Input parameter samples were generated by sampling the uniform distribution $U_[\\mu - \\sigma, \\mu + \\sigma]$, where $\\mu$ and $\\sigma$ are the mean and standard deviation of the parameter's posterior distribution respectively. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
# plot but with mean and sd instead rather than individual
uniform_sims_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none") +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

@@

```{r fig-sensitvity-pre-outbreak-abx, include=TRUE}
#| echo: false
#| fig-cap: "Sensitivity analysis of the model to changes in the pre-outbreak antimicrobial prescription rate. The y axis shows the total number of nosocomial infections during the outbreak period."
sens_abx_prescription_pre_outbreak_outbreak_cases %>% 
  group_by(`antibiotic-prescription-rate`) %>% 
  summarise(mean_cases = mean(`outbreak-cases`),
            sd_cases = sd(`outbreak-cases`)) %>% 
  ggplot(aes(x = `antibiotic-prescription-rate`, y = mean_cases)) +
  geom_line() + 
  # add sd
  geom_line(data = . %>% 
              mutate(upper = mean_cases + sd_cases,
                     lower = mean_cases - sd_cases),
            aes(y = upper), linetype = "dashed") +
  geom_line(data = . %>% 
              mutate(upper = mean_cases + sd_cases,
                     lower = mean_cases - sd_cases),
            aes(y = lower), linetype = "dashed") +
  xlab("Antimicrobial prescription rate") +
  ylab("Number of infected patients")
```

@@

>>>>

---
title: Gastro-intestinal transmission and colonisation of _Clostridioides difficile_ and antimicrobial-resistant bacteria in healthcare settings -- calibration of an agent-based model
authors:
  - name: Alessandro Gerada
    affiliation: University of Liverpool
    roles: writing
    corresponding: true
bibliography: references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/refs/heads/master/elsevier-vancouver.csl
filters:
  - acronyms
acronyms:
  keys:
  - shortname: CDI
    longname: _Clostridium difficile_ infection
    markdown: longname
  - shortname: VRE
    longname: vancomycin-resistant _Enterococcus_
    markdown: longname
  - shortname: ESBL
    longname: extended-spectrum beta-lactamase
  - shortname: CPE
    longname: carbapenemase-producing Enterobacterales
  - shortname: WHO
    longname: World Health Organization
  - shortname: IPC
    longname: infection prevention and control
  - shortname: ABM
    longname: agent-based model
  - shortname: MRSA
    longname: methicillin-resistant _Staphylococcus aureus_
    markdown: longname
  - shortname: SARS-CoV-2
    longname: severe acute respiratory syndrome coronavirus 2
  - shortname: ABC
    longname: approximate Bayesian computation
  - shortname: IQR
    longname: interquartile range
  - shortname: FIFO
    longname: first-in, first-out
  - shortname: SD
    longname: system dynamics
  insert_loa: false
---

@@

```{r fig-sd-model, include=TRUE}
#| echo: false
#| fig.cap: "System dynamics model. $S$ and $I$ are the number of susceptible and infected patients, respectively; $A$ and $D$ are the rates of admitted and discharged patients, respectively; $p$ is the community colonisation rate; $\\beta$ is the transmission coefficient; and $r$ is the rate of random (*de novo*) infection."
knitr::include_graphics("./images/sd_model.svg")
```

@@

A similar approach was used to generate outbreaks in the stochastic system dynamics model, with the difference that the parameters were changed at: 1) onset of outbreak; and 2) end of outbreak. The parameters that were changed were: 1) $\beta$ (transmission coefficient); and 2) $p$ (community colonisation rate). The parameter $r$ (rate of random infection) was kept constant throughout the simulation.

@@

```{r fig-screenshot, include=FALSE}
#| echo: false
#| out.height: "7in"
#| fig.cap: "Screenshot of model graphical user interface during a simulation run"
knitr::include_graphics("./images/screenshot.png")
```

@@

### Calibration of \acr[first_use=TRUE]{ABM}

Rejection sampling \acr{ABC} was employed to fit the \acr{ABM} to the observed outbreak data [@sunnakerApproximateBayesianComputation2013]. Priors were extracted from existing literature; flat priors when data were unavailable. A summary of the priors used is available in @tbl-priors. The proportion of toilets that were shared between multiple patients was set to 60%.

@@

As a posterior predictve check, the posterior parameter distributions were used to simulate `r length(unique(uniform_sims$siminputrow))` outbreak scenarios that were compared to the observed rates. To sample from the posterior distribution (i.e., the parameter sets retained after the rejection ABC algorithm) were samples with replacement were taken and used as an input to the stochastic \acr{SD} model.

@@

### Calibration of the \acr[first_use=TRUE]{SD} model

\acr[first_use=TRUE, case=sentence]{ABC} was also used to calibrate the stochastic \acr{SD} model to the observed outbreak data, although the calibration was performed in a single step that encompassed the entire period of the observed data (i.e., pre-outbreak, outbreak, and control periods).

@@

To assess the sensitivity of the calibrated \acr{ABM} to changes in parameters, a single-parameter local sensitivity analysis was performed [@railsback232SensitivityAnalysis2019]. Four parameters that reflected modifiable infection control practices were chosen for the sensitivity analysis: 1) antimicrobial prescription rates; 2) the mean delay between patient infection and transfer to a side room; 3) toilet cleaning rate; and 4) toilet cleaning effect. To conduct the sensitivity analysis, all the parameters except the one of interest were fixed at the posterior mean values, determined using the rejection \acr{ABC} Next, the parameter of interest was varied through a range of values, and used as an input to the simulation model at each value. The output metric was the number of cases of hospital-acquired infection/colonisation within the outbreak period. For each parameter value, the simulation was run with `r sens_n_seeds` different random number generator seeds to account for stochasticity in the model -- the output metric was the mean number of cases across the `r sens_n_seeds` runs.

@@

Both the \acr{ABM} and the \acr{SD} model were able to simulate the faecal-oral transmission of infection in a hospital setting. Although the \acr{SD} was computationally simpler, and produced simulated infection rates that were very similar to the observed rates (see @fig-posterior-predictive-check), the \acr{ABM} was based on a simulation that more closely resembled the workings of a real hospital. Importantly, the parameters of the \acr{ABM} were directly based on modifiable infection control practices, such as toilet cleaning frequency and antimicrobial prescription rates. Therefore, the main analyses and senstivity analyses were performed using the \acr{ABM}.

After calibrating the \acr{ABM} using \acr{ABC}, three groups of parameters were produced: 1) baseline parameters, which aimed to generate similar rates of infection to the basal rate (i.e., pre- and post-outbreak); 2) outbreak parameters, which led to the increased transmission observed during the outbreak period; and 3) control parameters, which reflect enhanced infection control practice that led to outbreak control. A summary of the posterior estimates of the parameters (reported as the mean and standard deviation of the parameters from the rejection \acr{ABC}) is shown in @tbl-posteriors. The densities of the posterior parameter distributions are shown in @fig-posterior-densities. In line with the prior bounds described in @sec-outbreak-generation, the posterior distributions of the parameters were consistent with the expected changes in the parameters during the outbreak and control periods (e.g., the toilet cleaning rate changed from `r b_toilet_clean_rate` to `r o_toilet_clean_rate`, and to `r c_toilet_clean_rate` during the baseline, outbreak, and control periods respectively).

@@

```{r fig-consts-sims-mean-sd, include=FALSE}
#| echo: false
#| fig-cap: "Simulated rates of _C. difficile_ infection generated by using the posterior parameter distributions. Input parameter samples were generated by sampling the uniform distribution $U_[\\mu - \\sigma, \\mu + \\sigma]$, where $\\mu$ and $\\sigma$ are the mean and standard deviation of the parameter's posterior distribution respectively. The red line shows the true observed infection rates. The black solid line shows the median simulated rates; the dashed lines show the 25^th^ and 75^th^ quantiles."
# plot but with mean and sd instead rather than individual
uniform_sims_rates %>% 
  pivot_wider(names_from = date_sim, values_from = rate) %>% 
  pivot_longer(cols = !c(`random-seed`, `siminputrow`), names_to = "date_sim", values_to = "rate") %>% 
  mutate(date_sim = ymd(date_sim)) %>% 
  group_by(date_sim) %>%
  summarise(m = median(rate),
            q_high = quantile(rate, 0.75),
            q_low = quantile(rate, 0.25),
            .groups = "keep") %>%
  ggplot() +
  geom_line(aes(x = date_sim, y = m)) +
  geom_line(aes(x = date_sim, y = q_high), linetype='dashed') +
  geom_line(aes(x = date_sim, y = q_low), linetype='dashed') + 
  geom_line(data = long_observed_ss, aes(x = x, y = rates), color = "red") +
  guides(color="none") +
  xlab("Date") + ylab("Rate (cases per 1000 patient days)")
```

@@

```{r fig-sensitvity-pre-outbreak-abx, include=FALSE}
#| echo: false
#| fig-cap: "Sensitivity analysis of the model to changes in the pre-outbreak antimicrobial prescription rate. The y axis shows the total number of nosocomial infections during the outbreak period."
sens_abx_prescription_pre_outbreak_outbreak_cases %>% 
  group_by(`antibiotic-prescription-rate`) %>% 
  summarise(mean_cases = mean(`outbreak-cases`),
            sd_cases = sd(`outbreak-cases`)) %>% 
  ggplot(aes(x = `antibiotic-prescription-rate`, y = mean_cases)) +
  geom_line() + 
  # add sd
  geom_line(data = . %>% 
              mutate(upper = mean_cases + sd_cases,
                     lower = mean_cases - sd_cases),
            aes(y = upper), linetype = "dashed") +
  geom_line(data = . %>% 
              mutate(upper = mean_cases + sd_cases,
                     lower = mean_cases - sd_cases),
            aes(y = lower), linetype = "dashed") +
  xlab("Antimicrobial prescription rate") +
  ylab("Number of infected patients")
```

@@

```{r}
# compare ABM community colonisation rate to SD model p
mean_sims %>% 
  summarise(mean = mean(`community-colonisation-rate`),
            sd = sd(`community-colonisation-rate`))
```
